Towards Detecting Stalkerware Through Network Traces

### 2019-03-22 paper/main.tex
Stalkerware ecosystem updates

Stalkerware is not as widely studied as other types of malware, and much less 
information is available regarding it. Although it is superficially 
similar to other types of malware, there exist a few key differences. First, 
stalkerware is meant to be installed 
deliberately, by someone with physical access to the device that is to be 
monitored. In the context of intimate partner abuse, this presents an enormous 
threat. An abuser will often know or be able to coerce passwords to the device 
from the victim, making traditional defense measures irrelevant. This is one 
example of how traditional assumptions about security often do not apply in the 
context of IPV (intimate partner violence). Second, the 
threat model used by malware developers includes sophisticated antivirus and 
intrusion detection systems (IDSs). Both malware and stalkerware need to hide 
their presence on the target device, but malware is much more likely to 
attempt to obfuscate its network traffic. Stalkerware does not: it does not 
need to. Most antivirus does not detect apps capable of monitoring a target as 
potentially dangerous \cite{chatterjee_spyware_2018}. This presents an 
opportunity to glean information about stalkerware that is not available in the 
study of other forms of malware. Third, stalkerware falls 
for the most part into two categories: apps that are explicitly advertised for 
surveilling spouses and partners, and apps that have a different stated use but 
can be repurposed to stalk a target. These are usually marketed for theft 
prevention, finding lost devices, parental control of children, and employer 
control of employees. Previous literature refers to these as "dual-use" 
apps\cite{chatterjee_spyware_2018}. Some dual-use apps are ones that the 
victim wants to keep installed, such as 
``find my phone'' apps or social media apps that leak the user's location. The 
most troubling example of this is Facebook, which most victims want to keep 
so that they can contact their support network for help.  Determining if a 
particular dual-use app is being used for intimate partner surveillance (IPS) 
presents an enormous challenge. Given the wide spectrum of apps that can be 
used to facilitate IPS and the limited scope of this project, we chose to focus 
on more overt spyware.

@@

We began by identifying a set of apps that appear to be 
marketed primarily for IPS. We focused exclusively on Android apps, since our 
two test phones were both Androids. Apps that appear in Google searches for 
terms like 
"catch a cheating spouse," use vague language that appear to be marketed 
towards IPS, and label themselves as "spyware" became candidates that we 
considered. Overt spyware is not generally available from the app store - it 
must be downloaded from the manufacturer sites. It also costs much more than a 
typical app, with monthly 
subscriptions ranging from \$12 to \$70 per month. The exception is an app that 
we call SpyToMobile, which survives on the app store by frequently 
``reskinning'' itself (see section \ref{reskinning}).  
We selected several of these to test. All of these apps brand themselves as 
spyware and advertise that their products are undetectable on the target phone, 
despite also having disclaimers in small text that their software should not be 
used for any illegal activity. Some use language that appears designed to 
appeal to abusers. For example, underneath one promotional 
video, the app TheOneSpy has the disclaimer: ``Note for this video: Don't 
confuse with the world `Loved Ones' as spouse or Girl friends! Word Loved ones 
has used for kids and teens.'' Many of these apps also refer to the tracked 
phone as the ``target'' phone, and advertise that once installed, the app is in 
``stealth mode,'' meaning completely undetectable. None of these required 
rooting the phone. When installed, they ask for all the permissions they 
require, including device administrator privileges. They are capable of 
blocking incoming calls, recording messages that have been deleted, recording 
the screen of the phone using accessibility features, monitoring apps such as 
Facebook, WhatsApp, Kik, and Viber, and more.

Other apps were overt spyware, but nevertheless available from the Play Store. 
They fell into two categories: spyware that reskinned itself to appear to be 
many different apps, and apps designed for parents to keep track of children 
(or in one case, for adult children to track elderly parents.) For the most 
part, this latter category was offered by antivirus companies such as Avast and 
Norton. These apps offered 
multiple features, such as call screening, blocking Google search results, 
recording text messages and call logs, and ``panic buttons'' that would allow 
the surveilled phone to contact the surveillant and raise an alarm. They do not 
hide their app icons and generally make an effort to notify the user that the 
monitoring app is installed on their phone. The spyware that did not notify 
users of its presence mostly consisted of one app, called SpyToMobile, that was 
provided on the Play Store under many different names (see section \ref{}).

We also identified many apps that were designed to track location or text 
messages, but were less comprehensive than the overt, off-store spyware. They 
track one feature each instead of tracking location, communication, and 
browsing history all at once. For the most part, these apps were free and 
available on the Google Play Store. Most of these were designed to keep track 
of the locations of the users' friends.

\section{The Stalkerware Ecosystem}
Some spyware has a notification that the app is running, but that can be 
disabled. Those ones don't hide their icon, but you might have to go to All 
Apps to find it.
\subsection{The Practice of ``Reskinning''}
\label{reskinning}

@@

Somewhere: put a note about the weird thing where the paid apps try to make you 
download stuff again. Maybe it's just us? Maybe they're trying to boost 
download count?

@@

The other category for stalkerware encompasses a large range of quality. Many 
apps advertised as stalkerware have absolutely no functionality other than 
advertisement of other apps. The few that are functional are often simplistic 
in both features and network patterns. Unlike the overt apps, however, the vast 
majority are usually free with advertisements.

@@

	Texting, Chat, Phone Spy by Maxlo & Domain:spy2mobile.com & SpyToMobile lookalike \\
\end{tabular}
\end{table*}

@@

\section{Wireshark Display Filters for Specific Signatures}
\begin{lstlisting}
ip.host contains dedicated.codero.net 
|| http.host contains hellospy
|| (ip.host contains data-api.spymyfone.com && ssl.handshake.extension.data 
contains "my.spymyfone.com") 
|| (ip.host contains ws.spyzie.com && ssl.handshake.extension.data contains 
"ws.spyzie.com") 
|| ip.host contains data-api.spymyfone.com 
|| ip.host contains ws.spyzie.com 
|| ip.host contains phonedata.me 
|| ssl.handshake.extension.data contains "phonedata.me" 
|| ip.host contains apiv4.alter757.info 
|| ssl.handshake.extension.data contains "thd.cc" 
|| ip.host contains thd.cc 
|| ip.host contains mspyonline.com 
|| ip.host contains corexchange.com && ssl.handshake.extensions_server_name 
contains messages02.smstracker.com 
|| ip.host contains life360.helpshift.com 
|| ip.host contains antitheft.ns1.ff.avast.com 
|| (ip.host contains production-finderapps && 
ssl.handshake.extensions_server_name contains "api.familysafetyproduction.com") 
|| ip.host contains production-finderapps 
|| ssl.handshake.extensions_server_name contains "celulocen.zendesk.com" 
|| ip.host contains www.zoemob.com 
|| ip.host contains unifiedlayer.com 
|| http.host matches "callsmstracker" 
|| ip.dst_host matches "onlinehome-server" 
|| (ip.dst_host matches "spoc-mu(|w|e)1.norton" && http.user_agent matches 
"(android|ios)")
|| ssl.handshake.extensions_server_name matches "family.norton" 
|| (udp && ip.host contains stun.l.google.com) 
|| http.host == relay.google.com && http.user_agent matches test 
|| http.host matches "97.logger.mobi" 
|| ssl.handshake.extensions_server_name matches "gpswox" 
|| ip.dst_host matches "gpswox" 
|| ssl.handshake.extensions_server_name matches "geozilla" 
|| ip.dst_host matches "clients.your-server.de" 
|| (ip.addr == 66.147.232.135 && tcp.port == 8080 && http.request.uri contains 
"signalr") 
|| ip.addr == 66.147.232.135 && tcp.port == 8080 
|| ssl.handshake.extensions_server_name matches "findmykids" 
|| ssl.handshake.extension.data contains "findmykids"  
||  ip.dst_host matches "(trackmyphones.com)" 
|| http.host matches "callsmstracker" 
|| ip.dst_host matches "onlinehome-server" 
\end{lstlisting}
\section{General Signature Wireshark Display Filters}

>>>>

Stalkerware is not as widely studied as other types of malware, and much less 
information about it is available. There are a few key differences between it 
and other forms of malware. First, stalkerware is meant to be installed 
by someone with physical access to the target device, not remotely by 
exploiting vulnerabilities. In the 
context of intimate partner abuse, this presents an enormous 
threat, given how frequently the abuser has unfettered access to the target 
phone. Second, the 
threat model used by malware developers includes sophisticated antivirus and 
intrusion detection systems (IDSs). Both malware and stalkerware need to hide 
their presence on the target device, but malware is much more likely to 
attempt to obfuscate its network traffic to thwart these adversaries. 
Stalkerware does not: it does not 
need to. Most antivirus does not detect apps capable of monitoring a target as 
potentially dangerous \cite{chatterjee_spyware_2018}. This presents an 
opportunity to glean information about stalkerware that is not available in the 
study of other forms of malware. At the same time, it limits a researcher's 
ability to use information collected by antivirus as a source of information 
about stalkerware. Third, stalkerware falls 
for the most part into two categories: apps that are explicitly advertised for 
surveilling spouses and partners, and apps that have a different stated use but 
can be repurposed to stalk a target. These are usually marketed for theft 
prevention, finding lost devices, parental control of children, and employer 
control of employees. Previous literature refers to these as "dual-use" 
apps\cite{chatterjee_spyware_2018}. Some dual-use apps are ones that the 
victim wants to keep installed, such as 
``find my phone'' apps or social media apps that leak the user's location. The 
most troubling example of this is Facebook, which most victims want to keep 
so that they can contact their support network for help.  Determining if a 
particular dual-use app is being used for intimate partner surveillance (IPS) 
presents an enormous challenge. Given the wide spectrum of apps that can be 
used to facilitate IPS and the limited scope of this project, we chose to focus 
on more overt spyware.

@@

We began by identifying a set of apps that appear to be 
marketed primarily for IPS. We focused exclusively on Android apps, since our 
two test phones were both Androids. Apps that appear in Google searches for 
terms like 
"catch a cheating spouse," use vague language that appear to be marketed 
towards IPS, and label themselves as "spyware" became candidates that we 
considered. Overt spyware is not generally available from the app store - it 
must be downloaded from the manufacturer sites. It also costs much more than a 
typical app, with monthly 
subscriptions ranging from \$12 to \$70 per month. The exception is an app that 
we call SpyToMobile, which survives on the app store by frequently 
``reskinning'' itself (see section \ref{reskinning}).  
We selected several of overt apps to test. All of these apps brand themselves 
as spyware and advertise that their products are undetectable on the target 
phone, despite also having disclaimers in small text that their software should 
not be used for any illegal activity. Some use language that appears designed 
to appeal to abusers. For example, underneath one promotional 
video, the app TheOneSpy has the disclaimer: ``Note for this video: Don't 
confuse with the world `Loved Ones' as spouse or Girl friends! Word Loved ones 
has used for kids and teens.''\cite{noauthor_theonespy_2013} All of these apps 
advertise that once they are installed, they are in ``stealth mode'' and are 
completely undetectable. This is not entirely the case - they appear in the 
list of all applications, but take care to name themselves something innocuous 
such as ``Data Backup'' or ``System Service.'' None of the overt apps that we 
examined required rooting the phone. When installed, they ask for all the 
permissions they 
require, including device administrator privileges. They are then capable of 
blocking incoming calls, recording messages that have been deleted, recording 
the screen of the phone using accessibility features, monitoring apps such as 
Facebook, WhatsApp, Kik, and Viber, and more.

Other apps were explicitly intended as spyware, but nevertheless available from 
the Play Store. They fell into two categories: spyware that reskinned itself to 
appear to be many different apps, and apps designed for parents to keep track 
of children (or in one case, for adult children to track elderly parents.) For 
the most part, this latter category was offered by antivirus companies such as 
Avast and Norton. These apps offered 
multiple features, such as call screening, blocking Google search results, 
recording text messages and call logs, and ``panic buttons'' that would allow 
the surveilled phone to contact the surveillant and raise an alarm. They do not 
hide their app icons and generally make an effort to notify the user that the 
monitoring app is installed on their phone. The spyware that we found on the 
Google Play store that did \textit{not} notify users of its presence consisted 
of one app, which we call SpyToMobile after the title on its loading screen, 
that was provided on the Play Store under many different names (see section 
\ref{reskinning}).

We also identified many apps that were designed to track location or text 
messages, but were less comprehensive than the overt spyware. They 
track one feature each instead of tracking location, communication, and 
browsing history all at once. For the most part, these apps were free and 
available on the Google Play Store. Most of them were designed to keep track 
of the locations of the users' friends. They tended to be ineffective - nearly 
all of the apps we found that did not work fell into this category. Most of 
these apps were unable or unwilling to hide their icons, and some provide an 
explicit notification that the app is in use, although this notification can be 
easily disabled in the Android notification settings.

\section{The Stalkerware Ecosystem}

We noted several interesting characteristics of the stalkerware ecosystem, some 
of which are beneficial to researchers attempting to identify stalkerware.

\subsection{The Practice of ``Reskinning''}
\label{reskinning}

@@

\subsection{Does Stalkerware Download Other Malware?}

Most free stalkerware appears to exist to serve advertisements. We only saw one 
instance of an app that came packaged with other malware, which in this case 
was a cryptominer for a cryptocurrency called Kin Coin.

We also observe a strange phenomenon among at least two of the paid, overt 
spyware apps. Both claimed that they had stopped working after two or three 
days even though they still appeared to be updating, and asked the user to 
click on a link to check if the app was still active. In one case, the URL of 
this link appeared to indicate that it would always claim the app had stopped 
working. The site would then request that the user re-download the app. We were 
confused as to why these apps would do this. 
The version of the app that was downloaded the second time appeared to be the 
same as the version downloaded first, and if the manufacturer wanted to 
download malware onto the phone, they could just do so with the first 
download. Nevertheless, our strongest explanation for this behavior is still 
that the false warnings were intended to convince users to download malware. We 
leave the task of determining if this was truly the case to future work.

@@

Free stalkerware apps vary widely in quality. Many simply don't work, or claim 
for some reason that they will only begin to work when the SIM card of the 
phone is replaced. Multiple apps that claim to track location instead just 
return the location associated with the phone's area code. A pair of others 
appeared to guess randomly - one reported that a device in San Diego was in 
Fawn Creek, Kansas, and another reported the same device was in Detroit, 
Michegan. Another common glitch was for spyware that records texts to only 
record one participant's side of the conversation. Their main purpose seems to 
be to serve 
advertisements - we did not 
catch any of the non-functional apps contacting domains associated with malware 
(except when the app itself was marked as malware).

@@

		Texting, Chat, Phone Spy by Maxlo & Domain:spy2mobile.com & SpyToMobile 
		lookalike \\
	\end{tabular}

@@

\section{Wireshark Display Filters for Specific Signatures}
\begin{lstlisting}
ip.host contains dedicated.codero.net 
|| http.host contains hellospy
|| (ip.host contains data-api.spymyfone.com && ssl.handshake.extension.data 
contains "my.spymyfone.com") 
|| (ip.host contains ws.spyzie.com && ssl.handshake.extension.data contains 
"ws.spyzie.com") 
|| ip.host contains data-api.spymyfone.com 
|| ip.host contains ws.spyzie.com 
|| ip.host contains phonedata.me 
|| ssl.handshake.extension.data contains "phonedata.me" 
|| ip.host contains apiv4.alter757.info 
|| ssl.handshake.extension.data contains "thd.cc" 
|| ip.host contains thd.cc 
|| ip.host contains mspyonline.com 
|| ip.host contains corexchange.com && ssl.handshake.extensions_server_name 
contains messages02.smstracker.com 
|| ip.host contains life360.helpshift.com 
|| ip.host contains antitheft.ns1.ff.avast.com 
|| (ip.host contains production-finderapps && 
ssl.handshake.extensions_server_name contains "api.familysafetyproduction.com") 
|| ip.host contains production-finderapps 
|| ssl.handshake.extensions_server_name contains "celulocen.zendesk.com" 
|| ip.host contains www.zoemob.com 
|| ip.host contains unifiedlayer.com 
|| http.host matches "callsmstracker" 
|| ip.dst_host matches "onlinehome-server" 
|| (ip.dst_host matches "spoc-mu(|w|e)1.norton" && http.user_agent matches 
"(android|ios)")
|| ssl.handshake.extensions_server_name matches "family.norton" 
|| (udp && ip.host contains stun.l.google.com) 
|| http.host == relay.google.com && http.user_agent matches test 
|| http.host matches "97.logger.mobi" 
|| ssl.handshake.extensions_server_name matches "gpswox" 
|| ip.dst_host matches "gpswox" 
|| ssl.handshake.extensions_server_name matches "geozilla" 
|| ip.dst_host matches "clients.your-server.de" 
|| (ip.addr == 66.147.232.135 && tcp.port == 8080 && http.request.uri contains 
"signalr") 
|| ip.addr == 66.147.232.135 && tcp.port == 8080 
|| ssl.handshake.extensions_server_name matches "findmykids" 
|| ssl.handshake.extension.data contains "findmykids"  
||  ip.dst_host matches "(trackmyphones.com)" 
|| http.host matches "callsmstracker" 
|| ip.dst_host matches "onlinehome-server" 
\end{lstlisting}

\section{General Signature Wireshark Display Filters}
